---
title: "Portfolio Project1"
author: "Huan Xu"
date: "2/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Ranking U.S. academic institutions by their open-source contributions

csrankings.org is a well-known site that ranks the CS department of U.S. academic institutions 
by their contributions to the academia. It considers factors like the total citation count from 
an institution's CS faculty body, and the number of power individuals who made major contributions 
to specific fields of Computer Sciences.
As the cradle of many impactful open-source projects, GitHub is a also a place for programmers
to make contributions and push the frontier of Computer Sciences. However, there is yet a report on
how much each U.S. academic institution contributed in this race.

To the end of building such a report, this portfolio project used Google Big Query to extract
history committer information whose email ends with ".edu", and ranks U.S. academic
institutions by their open-source contribution.

## Transform data into tidy format

Based on the observation that information about which institution an email belongs to is located
at the left of the ".edu" suffix. I used regular expression and tidyr's extract function to pull 
this information and put them into the new column, "institution".

```{r}
library(tidyr)
library(readr)
# data can also be downloaded here: https://drive.google.com/file/d/1xDClwr3HyI4BAlX9KWsHL2MWbM1BYxDT/view?usp=sharing
commits <- read_csv("https://dl.dropboxusercontent.com/s/61vrfnyc6ddijuc/bq-results-20220210-012020-2529sugofmkh.csv")
commits$institution <- extract(commits, email, into = c("institution"), "@(?:.*\\.)*(.*)\\.edu$")$institution
head(commits)
```
## Aggregate data by institution column

To rank institutions by total commits from committers, we need to first aggregate the data 
by institutions and use sum() to calculate the sum of commits from their committers.

```{r}
library(dplyr)
commits_by_institution <- commits %>% 
  group_by(institution) %>%
  summarize(sum_commit = sum(num_commit),
            num_committer = n(),
            commit_per_committer = sum_commit / num_committer) %>%
  drop_na() %>%
  slice_max(order_by = sum_commit, n = 50)
head(commits_by_institution)
```

## Plot 1: Rank institutions by the accumulated number of commits they made

About this graph: 
For this bar plot, I made 3 design choices to more clearly present the information.
1) I filtered for top 50 institutions ordered by the total number of commits to selectively display
more relevant information.
2) I ranked the y axis by the total number of commits to encode more information.
3) I filled each bar with colors related to the number of commit per committer to represent
how "active" a school is for open-source projects on GitHub.

Insight: 
The higher a school is on the list, the more total contribution it has made. 
Hence, we can conclude that schools like MIT, Stanford, Berkeley, UMich, and WISC 
are some of the most celebrated institutions throughout GitHub.

The darker a school's color is, the more active it is for open-source projects.
Hence, we can conclude that schools like Utath, WUSTL, PSC, and KU are some of the most
activce institutions throughout GitHub. (WISC is also very active!)

Unexpected things:
WISC is ranked higher than CMU. Score 1 for badgers.
```{r rank, echo=FALSE}
library(ggplot2)
commits_by_institution$group<-cut(commits_by_institution$commit_per_committer, 
                                  breaks = c(0,50,80,100,150,200,800))
ggplot(commits_by_institution) +
   geom_bar(
     aes(sum_commit, reorder(institution, sum_commit), fill = group),
     width = 1, stat = "identity"
   ) +
  scale_fill_brewer(palette = "YlOrRd") +
   scale_x_continuous(label = scales::label_number_si(), expand = c(0, 0, 0.1, 0.1)) + # remove scientific notation. scales::comma() is also useful.
   labs(x = "Accumulated GitHub commits", y = "Institution", fill = "Commit per committer") +
   theme(
     axis.text.y = element_text(size = 6),
     axis.ticks = element_blank(),
     panel.grid.major.y = element_blank()
   )
```

## Plot 2: Power committer from UW-Madison

About this graph: 
For this bar plot, I made 2 design choices to more clearly present the information.
1) I filtered for top 50 committers ordered by the total number of commits to selectively display
more relevant information.
2) I ranked the y axis by the number of commits to encode more information.

Insight: 
May I acknowledge the fictional GitHub Award winners to the following badgers: 
1st place, Curtis Rueden.
2nd place, Radu Serban.
3rd place, Jaime Frey.
ðŸ‘ðŸ‘ðŸ‘ðŸ‘

Unexpected things:
Huge respect to Curtis Rueden, who had 16,844 commits on GitHub, more than 20% of
total commits from WISC.
```{r wisc, echo=FALSE}
commiters_from_madison <- commits %>%
  filter(institution == 'wisc') %>%
  group_by(name) %>%
  summarise(num_commit = sum(num_commit)) %>%
  slice_max(order_by = num_commit, n = 50)

ggplot(commiters_from_madison) +
   geom_bar(
     aes(num_commit, reorder(name, num_commit)),
     width = 1, stat = "identity"
   ) +
   scale_x_continuous(label = scales::label_number_si(), expand = c(0, 0, 0.1, 0.1)) + # remove scientific notation. scales::comma() is also useful.
   labs(x = "Number of GitHub commits", y = "Name") +
   theme(
     axis.text.y = element_text(size = 6),
     axis.ticks = element_blank(),
     panel.grid.major.y = element_blank()
   )
```
